---
header-includes:
   - \usepackage{amsmath}
   - \usepackage{amsfonts}
output:
  html_document: default
---

\newcommand{\mb}[1]{\mathbf{#1}}
\newcommand{\var}{\text{var}}
\newcommand{\cov}{\text{cov}}

```{r setup-V1-P1-I2, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
set.seed(42) # NIPALS starts with a random vector
```

```{r, echo=FALSE, include=FALSE}
library(nipalsMCIA)
library(ggplot2)
library(ComplexHeatmap)
```

<!-- Loading NCI60 data and running nipals_multiblock()
    whenever running from individual Rmarkdowns -->
```{r child="Load-Shared-Vars.Rmd", include=FALSE}
```

### Visualizing the clustering of samples by factor scores
We may also be interested in the clustering of samples by their factors scores, to do this analyses we will the the XX function.
```{r, fig.align='center'}
gs_scores = mcia_results$global_scores
global_scores_heatmap_ComplexHeatmap(gs_scores)
```
From these heatmap we can see that samples cluster quite strongly by factor 1 signals (to be continued....want to have sample names to explain more if possible)