---
header-includes:
   - \usepackage{amsmath}
   - \usepackage{amsfonts}
output:
  html_document: default
---

\newcommand{\mb}[1]{\mathbf{#1}}
\newcommand{\var}{\text{var}}
\newcommand{\cov}{\text{cov}}

```{r setup-V1-P1-I2, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
set.seed(42) # NIPALS starts with a random vector
```

```{r, echo=FALSE, include=FALSE}
library(nipalsMCIA)
```

<!-- Loading NCI60 data and running nipals_multiblock()
    whenever running from individual Rmarkdowns -->
```{r child="Load-Shared-Vars.Rmd", include=FALSE}
```

### Visualizing the clustering of samples by factor scores
We may also be interested in the clustering of samples by their factors scores which can be done using `global_scores_heatmap()`:
```{r, fig.align='center'}
global_scores_heatmap(mcia_results, coloring = 1)
```
From these heatmap we can see that samples cluster quite strongly by factor 1 signals (to be continued....want to have sample names to explain more if possible)