---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>",
                      fig.path = "man/figures/README-", out.width = "100%")
```

# nipalsMCIA: Software to Compute Multi-Block Dimensionality Reduction

<!-- badges: start -->
[![BioC status](http://www.bioconductor.org/shields/build/release/bioc/nipalsMCIA.svg)](https://bioconductor.org/checkResults/release/bioc-LATEST/nipalsMCIA)
[![R-CMD-check](https://github.com/Muunraker/nipalsMCIA/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/Muunraker/nipalsMCIA/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

This package computes Multiple Co-Inertia Analysis (MCIA) on multi-block data
using the Nonlinear Iterative Partial Least Squares (NIPALS) method.

Features include:  

* Efficient computation of deflation and variance enabling embedding of
  high-volume (e.g. single-cell) datasets.  
* Functionality to perform out-of-sample embedding.
* Easy-to-adjust options for deflation and pre-processing
* Multiple visualization and analysis options for sample- and feature-level
  embedding results
* Streamlined and well-documented and supported code that is consistent with
  published theory to enable more efficient algorithm development and extension

**References**

* **Mattessich et al. (2024) nipalsMCIA: Flexible Multi-Block Dimensionality Reduction in R via Non-linear Iterative Partial Least Squares, bioRxiv 2024.06.07.597819 (https://doi.org/10.1101/2024.06.07.597819)**
* Mattessich (2022) A Review of Multi-Block Dimensionality Reduction via Multiple
Co-Inertia Analysis, M.S. Thesis, Dept. of Mathematics, Tufts University (http://hdl.handle.net/10427/CZ30Q6773)
* Meng et al. (2014) A multivariate approach to the integration of multi-omics
datasets, BMC Bioinformatics 2014(15) (https://doi.org/10.1186/1471-2105-15-162)
* Hanafi et al. (2011) Connections between multiple co-inertia analysis and
consensus principal component analysis, Chemometrics and Intelligent Laboratory Systems 106 (1) (https://doi.org/10.1016/j.chemolab.2010.05.010.)

## Installation

This package can be installed [via Bioconductor](https://bioconductor.org/packages/release/bioc/html/nipalsMCIA.html):

```{r installation-release, echo = TRUE, results = "hide", message = FALSE, eval = FALSE}
if (!require("BiocManager", quietly = TRUE))
  install.packages("BiocManager")

BiocManager::install("nipalsMCIA")
```

You can install the development version of nipalsMCIA from [GitHub](https://github.com/) with:

```{r installation-devel, echo = TRUE, results = "hide", message = FALSE, eval = FALSE}
# install.packages("devtools")
devtools::install_github("Muunraker/nipalsMCIA", ref = "devel",
                         force = TRUE, build_vignettes = TRUE)
```

## Basic Example

The package currently includes one test dataset: `data_blocks`. This is a list
of dataframes containing  observations of variables from three omics types
(mRNA, proteins, and micro RNA)  on 21 cancer cell lines from the NCI60 cancer
cell lines. The data file includes a `metadata` data frame containing the cancer
type associated with each cell line. 

```{r}
# load the package and set a seed for reproducibility
library(nipalsMCIA)
set.seed(42)
```

```{r load-data}
data(NCI60) # import data as "data_blocks" and metadata as "metadata_NCI60"

# examine the data and metadata
summary(data_blocks)
head(metadata_NCI60)
table(metadata_NCI60)
```

*Note: this dataset is reproduced from the [omicade4 package](https://www.bioconductor.org/packages/release/bioc/html/omicade4.html)
(Meng et. al., 2014). This package assumes all input datasets are in sample by
feature format.*

The main MCIA function can be called on `data_blocks` and optionally can include
`metadata_NCI60` for plot coloring by cancer type:

```{r call-mcia, fig.dim = c(8, 4), fig.align = "center", message = FALSE}
# to convert data_blocks into an MAE object we provide the simple_mae() function
data_blocks_mae <- simple_mae(data_blocks, row_format = "sample",
                              colData = metadata_NCI60)

mcia_results <- nipals_multiblock(data_blocks_mae = data_blocks_mae,
                                  col_preproc_method = "colprofile",
                                  num_PCs = 10, tol = 1e-12,
                                  color_col = "cancerType")
```

Here `num_PCs` is the dimension of the low-dimensional embedding of the data chosen by the user. 
