---
title: "Mini-Lecture: Overview of the NIPALS MCIA Pipeline"
author: "Max Mattessich, Joaquin Reyna, Edel Aron, Anna Konstorum"
date: "Compiled: `r format(Sys.time(), '%d %B %Y')`"
output: 
    rmdformats::readthedown
vignette: >
  %\VignetteIndexEntry{Mini-Lecture.Overview-of-the-NIPALS-MCIA-pipeline}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup-ml-pipeline, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

The Mini-Lecture, Brief outline of the math behind MCIA, covered the
mathematical aspects of MCIA. This second Mini-Lecture will cover the full
MCIA pipeline, with a focus on the standard preprocessing of multi-block data
and post-processing following MCIA. The target audience of this Mini-Lecture
are bioinformaticians or any other general practitioners.

```{r, echo = FALSE, out.width = "100%", out.length = "80%", fig.align = "center"}
knitr::include_graphics("images/MCIA-Pipeline-V5.svg")
```


## Session Info 

```{r session-info, echo = FALSE} 
sessionInfo()
```
