---
title: "Analysis of Multi-Omics Data with Multiple Co-Inertia Analysis via Nonlinear Iterative Partial Least Squares (NIPALS)"
author: "Maximilian Mattessich"
date: '2022-06-27'
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Quick Start 

The package can be installed directly from Github via the \verb+devtools+ package:

```{r installing package, echo = T, results = 'hide',error=FALSE, warning=FALSE,message=FALSE}
install.packages("devtools",repos = "http://cran.us.r-project.org") 
library(devtools)# required for install_github

install_github("Muunraker/NIPALS-MCIA")
library(nipalsMCIA)

```

A test dataset of NCI60 cancer cell line data with 3 blocks (mRNA, proteins, and micro-RNA) is included to illustrate how the main MCIA function works with multi-omics data:
```{r loading data}
data(NCI60) # loads data as data_blocks
summary(data_blocks)
```

The preprocessing function and main NIPALS iteration can then be applied to the data. The results include a plot of the first two global scores:
```{r computing MCIA,message=FALSE}

# Main NIPALS function
results <- nipals_multiblock(data_blocks, preprocMethod='colprofile',num_PCs = 10,
                             plots='global', tol=1e-12)

```
