---
title: "Vignette1 - Introduction & Main"
author:
    - Max Mattessich
    - Joaquin Reyna
    - Anna Konstorum
date: "9/1/2022"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, echo=FALSE}
library(nipalsMCIA)
```

## Part 1: Interpreting Global Scores
MCIA calculates a global scores matrix that is $n \times k$ where $n$ is the number of samples and $k$ is the number of factors. Each entry $M_{ik}$ represents the strength of factor $k$ to sample $i$. As such, you may be interested in further dissecting the global scores matrix in order to understand, for a given sample, what factors capture the most the MCIA decomposition. 
.
.
.
(tbd)

## Part 2: Interpreting Global Loadings
In addition to the global scores matrix, MCIA also calculates a global loadings matrix that is $(m_1 + ... + m_j + ... + m_R) \times k$ where $m_j$ is the number of features within the omics matrix $X^j$ and $K$ is the number of factors calculated. This second matrix provides information as to the contribution  


```{r}
data(NCI60)
names(data_blocks)
```


<!--
#```{r}
#results <- nipals_multiblock(data_blocks, preprocMethod='colprofile',num_PCs = 10,
#                             plots='global', tol=1e-12)
#feature_mat <- data_blocks[["mrna"]]
#latent_mat <- results$global_scores
#```

#```{r}
######### global_loadings_heatmap_ComplexHeatmap() ########
#global_loadings_heatmap_ComplexHeatmap(results$global_loadings,
#                                       data_blocks,
#                                       "mrna",
#                                       seq(1,10))
#```
-->






