---
title: "Vignette1 - Analysis of MCIA decomposition"
author:
    - Max Mattessich
    - Joaquin Reyna
    - Anna Konstorum
date: "9/1/2022"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(nipalsMCIA)
library(ggplot2)
```

## Part 1: Interpreting Global Factor Scores
MCIA calculates a global factor scores matrix that is $n \times k$ where $n$ is the number of samples and $k$ is the number of factors. Each entry $M_{ij}$ represents the strength of factor $j$ to sample $i$. As such, you may be interested in further dissecting the global scores matrix in order to understand, for a given sample, what factors capture the most the MCIA decomposition...(to be continued)

### Running MCIA on NCI60 Data and basic visualization
```{r}
data(NCI60)
results <- nipals_multiblock(data_blocks, preprocMethod='colprofile',num_PCs = 10, tol=1e-12)
```

### Visualize only global factor scores
```{r visualize training model, fig.height=4}
cancer_type <- substr(rownames(data_blocks$mrna), 1, 2)
MCIA_out<-data.frame(results$global_scores[,1:2])
MCIA_out$Cancer_type<-cancer_type
colnames(MCIA_out)<-c("Factor_1", "Factor_2", "Cancer_type")
ggplot(data = MCIA_out, aes(x=Factor_1, y=Factor_2, color=Cancer_type))+
  geom_point(size=3) + 
  theme_bw() +
  ggtitle("MCIA for NCI60 training data")
```

### Visualizing the clustering of samples by factor scores
```{r}
gs_scores = results$global_scores
colnames(gs_scores) = paste0('F', seq(1, ncol(results$global_scores)))
p = ComplexHeatmap::Heatmap(gs_scores, 
    name = "GS Score", 
    column_title = "Factors",
    row_title = "Samples",
    row_names_gp = grid::gpar(fontsize = 7),
    show_column_names = T,
    show_row_names = T,
    row_names_side = "right"
)    
p
```

## Part 2: Interpreting Global Loadings
In addition to the global scores matrix, MCIA also calculates a global loadings matrix that is $(m_1 + ... + m_j + ... + m_R) \times k$ where $m_j$ is the number of features within the omics matrix $X^j$ and $K$ is the number of factors calculated. This second matrix provides information as to the contribution  

### Pseudoeigenvalues representing the contribution of each omic to the global factor score
...(to be continued)

### Scree Plot: Visualizing the top features per factor
...(to be continued)

### Pathway analysis for the top factors using data from gene-centric omics blocks 
...(to be continued)
