---
title: "Predicting New MCIA scores"
author: "Max Mattessich, Joaquin Reyna, Edel Aron, Anna Konstorum"
date: "Compiled: `r format(Sys.time(), '%d %B %Y')`"
output: 
    rmdformats::readthedown:
      toc_depth: 4
vignette: >
  %\VignetteIndexEntry{Predicting-New-Scores}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup-v2, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Predicting MCIA global (factor) scores for new test samples

It may be of interest to use the embedding that is calculated on a training
sample set to predict scores on a test set (or, equivalently, on new data).  

After loading the nipalsMCIA library (see the X vignette for instructions on
how to install), we randomly split the NCI60 cancer cell line data into
training and test sets.

```{r load-packages}
library(ggplot2)
library(nipalsMCIA)
```

```{r split-data}
data(NCI60)
set.seed(8)

num_samples <- dim(data_blocks[[1]])[1]
num_train <- round(num_samples * 0.7, 0)
train_samples <- sample.int(num_samples, num_train)

data_blocks_train <- data_blocks
data_blocks_test <- data_blocks
for (i in seq_along(data_blocks)) {
  data_blocks_train[[i]] <- data_blocks_train[[i]][train_samples, ]
  data_blocks_test[[i]] <- data_blocks_test[[i]][-train_samples, ]
}
```

### Run NIPALS-MCIA on training data

```{r computing-MCIA-on-training-data, message = FALSE, fig.show = "hide"}
MCIA_train <- nipals_multiblock(data_blocks_train,
                                preproc_method = "colprofile", num_PCs = 10,
                                plots = "none", tol = 1e-9)
```

### Visualize model on training data using metadata on cancer type

```{r visualize-training-model, fig.height = 2.5, fig.align = 'center'}
MCIA_train$metadata <- data.frame(metadata_NCI60[train_samples, ],
                                  row.names = rownames(data_blocks_train$mrna))
colnames(MCIA_train$metadata) <- c("cancerType")
meta_colors <- get_metadata_colors(MCIA_train, 1)

MCIA_out <- data.frame(MCIA_train$global_scores[, 1:2])
MCIA_out$cancerType <- MCIA_train$metadata[, 1]
colnames(MCIA_out) <- c("Factor.1", "Factor.2", "cancerType")

ggplot(data = MCIA_out, aes(x = Factor.1, y = Factor.2, color = cancerType)) +
  geom_point(size = 3) +
  theme_bw() +
  scale_color_manual(values = meta_colors) +
  ggtitle("MCIA for NCI60 training data") 
```

### Generate factor scores for test data using the MCIA_train model

We use the \verb+predict_gs+ function to generate new factor scores on the test
data set using the MCIA_train model above

```{r generate-new-scores, message = FALSE}
MCIA_test_scores <- predict_gs(MCIA_train, data_blocks_test)
```

### Visualize new scores with old

We once again plot the top two factor scores for both the training and test
datasets

```{r visualize-both-models, fig.height = 2.5, fig.align = 'center'}
MCIA_out_test <- data.frame(MCIA_test_scores[, 1:2])
MCIA_out_test$cancerType <- metadata_NCI60[-train_samples, ]

colnames(MCIA_out_test) <- c("Factor.1", "Factor.2", "cancerType")
MCIA_out_test$set <- "test"
MCIA_out$set <- "train"
MCIA_out_full <- rbind(MCIA_out, MCIA_out_test)
rownames(MCIA_out_full) <- NULL

ggplot(data = MCIA_out_full, aes(x = Factor.1, y = Factor.2,
                                 color = cancerType, shape = set)) +
  geom_point(size = 3) +
  theme_bw() +
  scale_color_manual(values = meta_colors) +
  ggtitle("MCIA for NCI60 training and test data")
```

## Session Info 

```{r session-info, echo = FALSE} 
sessionInfo()
```